Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-13 _tex/paper2.tex
Built site for gh-pages

Building on the arguments developed in Section~\ref{sec-theory-issue2}
and the principles of Structural Equation Modeling (SEM)
\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
the conceptual-population model then includes two hierarchical
structural components. Each component defines how different observed or
latent variables influence the primary latent variable of interest
\citep{Everitt_et_al_2010}. This addition facilitates analyses that
integrate the trait measurement and hypothesis testing processes,
simultaneously accounting for the traits and their associated
uncertainties, thereby reducing bias and enhancing the measurement
precision.

Figure~\ref{fig-cj09} displays the two hierarchical structural
components in the motivating example. Specifically, the top branch of
DAG~\ref{fig-cj09_dag} depicts the first component. This component
describes how the students' writing-quality trait \((T_{I})\) and a set
of \emph{relevant} text-related variables \(X_{IA}\) causally influence
the texts' written-quality trait \(T_{IA}\). The DAG also highlights the
idiosyncratic errors \(e_{IA}\), which capture variations in the texts'
trait that neither the students' trait nor other relevant text-related
variables can explain. Here, \(T_{I}\) and \(e_{IA}\) are latent vectors
of length \(n_{I}\) and \((n_{I} \cdot n_{A})\), respectively.
Additionally, \(X_{IA}\) is an observed matrix with a dimension of
\((n_{I} \cdot n_{A})\) rows and \(p_{IA}\) columns.

@@

A cautious reader will notice that these descriptions rely on several
relevant variables. In this study, a \emph{relevant} or a
\emph{sufficient adjustment} set of variables (potentially an empty set)
refers to those that satisfy the \emph{backdoor criterion}
\citep[pp.~37]{Neal_2020}, that is, variables that block all non-causal
paths without introducing new ones
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.

@@

>>>>

Building on the arguments developed in Section~\ref{sec-theory-issue2}
and the principles of Structural Equation Modeling (SEM)
\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
the conceptual-population model includes two hierarchical structural
components. Each component defines how different observed or latent
variables influence the primary latent variable of interest
\citep{Everitt_et_al_2010}. This addition facilitates analyses that
integrate the trait measurement and hypothesis testing processes,
simultaneously accounting for the traits and their associated
uncertainties, thereby reducing bias and enhancing the measurement
precision.

Figure~\ref{fig-cj09} displays the two hierarchical structural
components for the motivating example. Specifically, the top branch of
DAG~\ref{fig-cj09_dag} depicts the first component. This component
describes how the students' writing-quality trait \(T_{I}\) and a set of
\emph{relevant} text-related variables \(X_{IA}\) causally influence the
texts' written-quality trait \(T_{IA}\). The DAG also highlights the
idiosyncratic errors \(e_{IA}\), which capture variations in the texts'
trait that neither the students' trait nor other relevant text-related
variables can explain. Here, \(T_{I}\) and \(e_{IA}\) are latent vectors
of length \(n_{I}\) and \((n_{I} \cdot n_{A})\), respectively.
Additionally, \(X_{IA}\) is an observed matrix with a dimension of
\((n_{I} \cdot n_{A})\) rows and \(p_{IA}\) columns.

@@

@@

A cautious reader will notice that the model descriptions of
Section~\ref{sec-theory-theoretical_P} rely on several relevant
variables. In this study, a \emph{relevant} or a \emph{sufficient
adjustment} set of variables (potentially an empty set) refers to those
that satisfy the \emph{backdoor criterion} \citep[pp.~37]{Neal_2020},
that is, variables that block all non-causal paths without introducing
new ones \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.
